cubble An R Package for Structuring Spatio-temporal Data

### 2022-04-17 arxiv/cubble-arxiv.tex
arxiv version

\documentclass[
]{jss}

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

\usepackage[utf8]{inputenc}

\author{
H. Sherry Zhang\\Monash University \And Dianne Cook\\Monash University
\AND Ursula Laa\\University of Natural Resources and Life Sciences
\AND Nicolas Langrené\\BNU-HKBU United International College
\AND Patricia Menéndez\\Monash University \AND
}
\title{\pkg{cubble}: An R Package for Structuring Spatio-temporal Data}

\Plainauthor{H. Sherry Zhang, Dianne Cook, Ursula Laa, Nicolas
Langrené, Patricia Menéndez}
\Plaintitle{cubble: An R Package for Structuring Spatio-temporal Data}

\Abstract{
Spatio-temporal data refer to measurements taken across space and time.
In practice, spatio-temporal data can be decomposed into a spatial and
temporal component: at one time, we would select a spatial location and
inspect the temporal trend; at other time, we might select one or
multiple time value(s) and explore the spatial distribution. Ideally, we
could make multiple maps and multiple time series to explore these
together, however, doing all of these actions is complicated when data
arrive fragmented in multiple objects. To make it easy to do all these
tasks, ideally spatial and temporal variables are in a single data
object that we can slice and dice in different ways to conduct different
visualisations. This work suggests a new data structure, \code{cubble},
that uses a nested form and a long form to organise spatial and temporal
variables in a single data object. The new data structure ensures that
data in the two forms are synchronised while been flexible to explore
the two aspect of spatio-temporal data. It also provides capability for
handling data with hierarchical structure, matching data from multiple
sources, constructing interactive graphics, and performing
spatio-temporal transformation. The paper will demonstrate \pkg{cubble}
with examples using Australian climate weather stations, river level
data, and climate reanalysis (ERA5) data.
}

\Keywords{spatial, temporal, spatio
temporal, \proglang{R}, environmental data, exploratory data analysis}
\Plainkeywords{spatial, temporal, spatio temporal, R, environmental
data, exploratory data analysis}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    H. Sherry Zhang\\
    Monash University\\
    21 Chancellors Walk, Clayton VIC 3800 Australia\\
  E-mail: \email{huize.zhang@monash.edu}\\

      Dianne Cook\\
    Monash University\\
    21 Chancellors Walk, Clayton VIC 3800 Australia\\
  E-mail: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}\\

      Ursula Laa\\
    University of Natural Resources and Life Sciences\\
    Gregor-Mendel-Straße 33, 1180 Wien, Austria\\
  E-mail: \href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}\\

      Nicolas Langrené\\
    BNU-HKBU United International College\\
    2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China\\
  E-mail: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}\\

      Patricia Menéndez\\
    Monash University\\
    21 Chancellors Walk, Clayton VIC 3800 Australia\\
  E-mail: \href{mailto:patricia.menendez@monash.edu}{\nolinkurl{patricia.menendez@monash.edu}}\\

@@

\usepackage{amsmath} \usepackage{array} \usepackage{xcolor}

\begin{document}

@@

\begin{CodeChunk}
\begin{figure}

@@

\caption[An illustration of different spatio-temporal data layouts]{An illustration of different spatio-temporal data layouts: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.}\label{fig:illu-input}
\end{figure}
\end{CodeChunk}

@@

In the \proglang{R} community, many data objects have been proposed to
work with specific spatio-temporal tasks, while few can freely explore
the data. \pkg{spacetime} \citep{spacetime} is an attempt for such a
data object, but the spatial and temporal classes it is built on have
now been gradually replaced by more recent implementations. The
\pkg{stars} \citep{stars} package uses an array to wrangle
spatio-temporal data, but the dense array it uses can quickly exhaust
the memory for data in the case of irregular spatial grid. While there
has not yet been a data object that makes it easy to wrangle
spatio-temporal data, there is a strong need for such tools given the
ubiquitous spatial and temporal data collected in the society, i.e.,
climate observations from weather stations, air quality variables from
air monitoring stations. The demand for such tools can also come from
spatial data analysts wishing to incorporate temporal data in their
analysis or vice versa.

@@

Multivariate spatio-temporal data can be conceptualised using a cubical
data model with reference to variables, time and space. This naturally
motivates the usage of array to represent and store spatio-temporal
data, as evident by satellite imageries, large climate models, among
others. The work by \citet{lu_multidimensional_2018} has discussed the
suitability of this representation through array algebra and practical
spatio-temporal tasks.

The cubical conceptual framework provides a generalisation to the
operation can be done on the data for visualisation. A paper by
\citet{bach_review_2014} has reviewed the temporal data visualisation
based on space-time cube operations. Notice that the term space-time
cube in their article ``does not need to involve spatial data'', but
referring to ``an abstract 2D substrate that is used to visualize data
at a specific time''. Despite its main focus is on temporal data, the
mindset of abstracting out data representation to construct
visualisation is applicable to our spatio-temporal data manipulation and
visualisation.

@@

To represent spatio-temporal data, \citet{spacetime} proposes four
layouts: full grid layout (STF), sparse grid layout (STS), irregular
layout (STI), and trajectory (STT). These layouts are implemented in the
\pkg{spacetime} package with underlying class \pkg{sp} \citep{sp} and
\pkg{xts} \citep{xts} for handling spatial and temporal variables. While
these layouts and classes are still being used and well-maintained
today, they are not compatible with the \pkg{tidyverse} ecosystem
\citep{tidyverse}, which has a huge influence on data analysis in
\proglang{R}, including in the spatial and temporal communities. The
package \pkg{sf} \citep{sf} uses data frame objects to represent spatial
data with an additional \code{geometry} column dedicating to feature
geometries \citep{jr_herring_opengis_2011} (e.g.~POINT, MULTIPOINT,
POLYGON, MULTIPOLYGON etc). For temporal data, \pkg{tsibble}
\citep{tsibble} builds on top of the \pkg{tibble} \citep{tibble} object
and explicitly includes date and time in columns as variables, rather
than data attributes.

Underlying tidyverse is the concept of tidy data \citep{tidydata}, which
considers how variables should be structured for analysis. In the three
formats (time-wide, space-wide, and long) summarised in the
\citet{spacetime} paper, only the long form is considered as tidy. The
main advantages of it over the time-wide and space-wide format is that
the latter treat either time or location as variable names (Section 3.1
in \citet{tidydata}) rather than values of variables, making them
difficult to be wrangled. One problem with the long format is that it
can be inefficient to store feature geometries, especially for large
MULTIPOLYGON over long time period. This motivates a new tidyverse
compatible data object to work with spatio-temporal data.

@@

\caption{An illustration of the nested and long form in cubble. The nested form defines each site in a row and nests the time varying variables into a single column \code{ts}. The long form cubble uses \code{id} and \code{t} to identify each row and stores the time invariant variables as an attribute, \code{spatial}.}\label{fig:illu-cubble}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_nested <- cubble::climate_flat |>
+   as_cubble(key = id, index = date, coords = c(long, lat))
R> cubble_nested
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble [366 x 4]>
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_long <- cubble_nested |> face_temporal()
R> cubble_long
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.97, -32.94, 133.55, -12.42]
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3
2 ASN00009021 2020-01-02     0  24.9  16.4
3 ASN00009021 2020-01-03     6  23.2  13  
4 ASN00009021 2020-01-04     0  28.4  12.4
5 ASN00009021 2020-01-05     0  35.3  11.6
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> attr(cubble_long, "spatial")
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 5 x 6
  id            lat  long  elev name           wmo_id
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl>
1 ASN00009021 -31.9  116.  15.4 perth airport   94610
2 ASN00010311 -31.9  117. 179   york            94623
3 ASN00010614 -32.9  117. 338   narrogin        94627
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120
5 ASN00015131 -17.6  134. 220   elliott         94236
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_back <- cubble_long |> face_spatial()
R> cubble_back
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble [366 x 4]>
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_mig <- cubble_long |> unfold(long, lat)
R> cubble_mig
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.97, -32.94, 133.55, -12.42]
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]
  id          date        prcp  tmax  tmin  long   lat
  <chr>       <date>     <dbl> <dbl> <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3  116. -31.9
2 ASN00009021 2020-01-02     0  24.9  16.4  116. -31.9
3 ASN00009021 2020-01-03     6  23.2  13    116. -31.9
4 ASN00009021 2020-01-04     0  28.4  12.4  116. -31.9
5 ASN00009021 2020-01-05     0  35.3  11.6  116. -31.9
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_mig |> face_spatial() |> face_temporal()
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.97, -32.94, 133.55, -12.42]
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3
2 ASN00009021 2020-01-02     0  24.9  16.4
3 ASN00009021 2020-01-03     6  23.2  13  
4 ASN00009021 2020-01-04     0  28.4  12.4
5 ASN00009021 2020-01-05     0  35.3  11.6
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # example with a tsibble created from climate_flat
R> raw <- climate_flat |>
+   tsibble::as_tsibble(key = id, index = date)
R> 
R> dt <-  raw |>
+   cubble::as_cubble(coords = c(long, lat))
R> dt
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_ts [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_ts [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_ts [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_ts [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_ts [366 x 4]>
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # add station-based features in the nested form.
R> dt |>
+   mutate(fabletools::features(ts, tmax, list(tmax_mean = mean, tmax_var = var)))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name          wmo_id ts       tmax_mean tmax_var
  <chr>       <dbl> <dbl> <dbl> <chr>          <dbl> <list>       <dbl>    <dbl>
1 ASN00009021 -31.9  116.  15.4 perth airport  94610 <tbl_ts>      25.7    38.6 
2 ASN00010311 -31.9  117. 179   york           94623 <tbl_ts>      26.2    51.1 
3 ASN00010614 -32.9  117. 338   narrogin       94627 <tbl_ts>      23.7    45.4 
4 ASN00014015 -12.4  131.  30.4 darwin airpo~  94120 <tbl_ts>      33.1     3.02
5 ASN00015131 -17.6  134. 220   elliott        94236 <tbl_ts>      34.6    24.7 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # read in the .nc file as a ncdf4 class
R> raw <- ncdf4::nc_open(here::here("data/era5-pressure.nc"))
R> 
R> # convert the variable q and z in the ncdf4 into a cubble
R> dt <- as_cubble(raw, vars = c("q", "z"))
\end{CodeInput}
\end{CodeChunk}

>>>>

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

\title{cubble An R Package for Structuring Spatio-temporal Data}

\author{
    H. Sherry Zhang
   \\
    Monash University \\
  21 Chancellors Walk, Clayton VIC 3800 Australia \\
  \texttt{\href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}} \\
   \And
    Dianne Cook
   \\
    Monash University \\
  21 Chancellors Walk, Clayton VIC 3800 Australia \\
  \texttt{\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}} \\
   \And
    Ursula Laa
   \\
    University of Natural Resources and Life Sciences \\
  Gregor-Mendel-Straße 33, 1180 Wien, Austria \\
  \texttt{\href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}} \\
   \And
    Nicolas Langrené
   \\
    BNU-HKBU United International College \\
  2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China \\
  \texttt{\href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}} \\
   \And
    Patricia Menéndez
   \\
    Monash University \\
  21 Chancellors Walk, Clayton VIC 3800 Australia \\
  \texttt{\href{mailto:patricia.menendez@monash.edu}{\nolinkurl{patricia.menendez@monash.edu}}} \\
  }

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\begin{document}
\maketitle

\begin{abstract}
Spatio-temporal data refer to measurements taken across space and time.
In practice, spatio-temporal data can be decomposed into a spatial and
temporal component: at one time, we would select a spatial location and
inspect the temporal trend; at other time, we might select one or
multiple time value(s) and explore the spatial distribution. Ideally, we
could make multiple maps and multiple time series to explore these
together, however, doing all of these actions is complicated when data
arrive fragmented in multiple objects. To make it easy to do all these
tasks, ideally spatial and temporal variables are in a single data
object that we can slice and dice in different ways to conduct different
visualisations. This work suggests a new data structure, \code{cubble},
that uses a nested form and a long form to organise spatial and temporal
variables in a single data object. The new data structure ensures that
data in the two forms are synchronised while been flexible to explore
the two aspect of spatio-temporal data. It also provides capability for
handling data with hierarchical structure, matching data from multiple
sources, constructing interactive graphics, and performing
spatio-temporal transformation. The paper will demonstrate \pkg{cubble}
with examples using Australian climate weather stations, river level
data, and climate reanalysis (ERA5) data.
\end{abstract}

\keywords{
    spatial, temporal, spatio temporal, R, environmental data,
    exploratory data analysis
  }

@@

\begin{figure}

@@

\caption{An illustration of different spatio-temporal data layouts: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.}\label{fig:illu-input}
\end{figure}

@@

In the \proglang{R} community, many data objects have been proposed to
work with specific spatio-temporal tasks, while few can freely explore
the data. \pkg{spacetime} (\protect\hyperlink{ref-spacetime}{E. Pebesma
2012}) is an attempt for such a data object, but the spatial and
temporal classes it is built on have now been gradually replaced by more
recent implementations. The \pkg{stars}
(\protect\hyperlink{ref-stars}{E. Pebesma 2021}) package uses an array
to wrangle spatio-temporal data, but the dense array it uses can quickly
exhaust the memory for data in the case of irregular spatial grid. While
there has not yet been a data object that makes it easy to wrangle
spatio-temporal data, there is a strong need for such tools given the
ubiquitous spatial and temporal data collected in the society, i.e.,
climate observations from weather stations, air quality variables from
air monitoring stations. The demand for such tools can also come from
spatial data analysts wishing to incorporate temporal data in their
analysis or vice versa.

@@

Multivariate spatio-temporal data can be conceptualised using a cubical
data model with reference to variables, time and space. This naturally
motivates the usage of array to represent and store spatio-temporal
data, as evident by satellite imageries, large climate models, among
others. The work by Lu, Appel, and Pebesma
(\protect\hyperlink{ref-lu_multidimensional_2018}{2018}) has discussed
the suitability of this representation through array algebra and
practical spatio-temporal tasks.

The cubical conceptual framework provides a generalisation to the
operation can be done on the data for visualisation. A paper by Bach et
al. (\protect\hyperlink{ref-bach_review_2014}{2014}) has reviewed the
temporal data visualisation based on space-time cube operations. Notice
that the term space-time cube in their article ``does not need to
involve spatial data'', but referring to ``an abstract 2D substrate that
is used to visualize data at a specific time''. Despite its main focus
is on temporal data, the mindset of abstracting out data representation
to construct visualisation is applicable to our spatio-temporal data
manipulation and visualisation.

@@

To represent spatio-temporal data, E. Pebesma
(\protect\hyperlink{ref-spacetime}{2012}) proposes four layouts: full
grid layout (STF), sparse grid layout (STS), irregular layout (STI), and
trajectory (STT). These layouts are implemented in the \pkg{spacetime}
package with underlying class \pkg{sp} (\protect\hyperlink{ref-sp}{E.
Pebesma and Bivand 2005}) and \pkg{xts}
(\protect\hyperlink{ref-xts}{Ryan and Ulrich 2020}) for handling spatial
and temporal variables. While these layouts and classes are still being
used and well-maintained today, they are not compatible with the
\pkg{tidyverse} ecosystem (\protect\hyperlink{ref-tidyverse}{Wickham et
al. 2019}), which has a huge influence on data analysis in \proglang{R},
including in the spatial and temporal communities. The package \pkg{sf}
(\protect\hyperlink{ref-sf}{E. J. Pebesma 2018}) uses data frame objects
to represent spatial data with an additional \code{geometry} column
dedicating to feature geometries
(\protect\hyperlink{ref-jr_herring_opengis_2011}{J.R. Herring 2011})
(e.g.~POINT, MULTIPOINT, POLYGON, MULTIPOLYGON etc). For temporal data,
\pkg{tsibble} (\protect\hyperlink{ref-tsibble}{Wang, Cook, and Hyndman
2020a}) builds on top of the \pkg{tibble}
(\protect\hyperlink{ref-tibble}{Müller and Wickham 2021}) object and
explicitly includes date and time in columns as variables, rather than
data attributes.

Underlying tidyverse is the concept of tidy data
(\protect\hyperlink{ref-tidydata}{Wickham 2014}), which considers how
variables should be structured for analysis. In the three formats
(time-wide, space-wide, and long) summarised in the E. Pebesma
(\protect\hyperlink{ref-spacetime}{2012}) paper, only the long form is
considered as tidy. The main advantages of it over the time-wide and
space-wide format is that the latter treat either time or location as
variable names (Section 3.1 in Wickham
(\protect\hyperlink{ref-tidydata}{2014})) rather than values of
variables, making them difficult to be wrangled. One problem with the
long format is that it can be inefficient to store feature geometries,
especially for large MULTIPOLYGON over long time period. This motivates
a new tidyverse compatible data object to work with spatio-temporal
data.

@@

\caption{An illustration of the nested and long form in cubble. The nested form defines each site in a row and nests the time varying variables into a single column \code{ts}. The long form cubble uses \code{id} and \code{t} to identify each row and stores the time invariant variables as an attribute, \code{spatial}.}\label{fig:illu-cubble}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_nested }\OtherTok{\textless{}{-}}\NormalTok{ cubble}\SpecialCharTok{::}\NormalTok{climate\_flat }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date, }\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}
\NormalTok{cubble\_nested}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # cubble:   id [5]: nested form
## # bbox:     [115.97, -32.94, 133.55, -12.42]
## # temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
##   id            lat  long  elev name           wmo_id ts                
##   <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
## 1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble [366 x 4]>
## 2 ASN00010311 -31.9  117. 179   york            94623 <tibble [366 x 4]>
## 3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble [366 x 4]>
## 4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble [366 x 4]>
## 5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble [366 x 4]>
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_long }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_nested }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_temporal}\NormalTok{()}
\NormalTok{cubble\_long}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # cubble:  date, id [5]: long form
## # bbox:    [115.97, -32.94, 133.55, -12.42]
## # spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]
##   id          date        prcp  tmax  tmin
##   <chr>       <date>     <dbl> <dbl> <dbl>
## 1 ASN00009021 2020-01-01     0  31.9  15.3
## 2 ASN00009021 2020-01-02     0  24.9  16.4
## 3 ASN00009021 2020-01-03     6  23.2  13  
## 4 ASN00009021 2020-01-04     0  28.4  12.4
## 5 ASN00009021 2020-01-05     0  35.3  11.6
## # ... with 1,825 more rows
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{attr}\NormalTok{(cubble\_long, }\StringTok{"spatial"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 6
##   id            lat  long  elev name           wmo_id
##   <chr>       <dbl> <dbl> <dbl> <chr>           <dbl>
## 1 ASN00009021 -31.9  116.  15.4 perth airport   94610
## 2 ASN00010311 -31.9  117. 179   york            94623
## 3 ASN00010614 -32.9  117. 338   narrogin        94627
## 4 ASN00014015 -12.4  131.  30.4 darwin airport  94120
## 5 ASN00015131 -17.6  134. 220   elliott         94236
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_back }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_long }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_spatial}\NormalTok{()}
\NormalTok{cubble\_back}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # cubble:   id [5]: nested form
## # bbox:     [115.97, -32.94, 133.55, -12.42]
## # temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
##   id            lat  long  elev name           wmo_id ts                
##   <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
## 1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble [366 x 4]>
## 2 ASN00010311 -31.9  117. 179   york            94623 <tibble [366 x 4]>
## 3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble [366 x 4]>
## 4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble [366 x 4]>
## 5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble [366 x 4]>
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_mig }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_long }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{unfold}\NormalTok{(long, lat)}
\NormalTok{cubble\_mig}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # cubble:  date, id [5]: long form
## # bbox:    [115.97, -32.94, 133.55, -12.42]
## # spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]
##   id          date        prcp  tmax  tmin  long   lat
##   <chr>       <date>     <dbl> <dbl> <dbl> <dbl> <dbl>
## 1 ASN00009021 2020-01-01     0  31.9  15.3  116. -31.9
## 2 ASN00009021 2020-01-02     0  24.9  16.4  116. -31.9
## 3 ASN00009021 2020-01-03     6  23.2  13    116. -31.9
## 4 ASN00009021 2020-01-04     0  28.4  12.4  116. -31.9
## 5 ASN00009021 2020-01-05     0  35.3  11.6  116. -31.9
## # ... with 1,825 more rows
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_mig }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_spatial}\NormalTok{() }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_temporal}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # cubble:  date, id [5]: long form
## # bbox:    [115.97, -32.94, 133.55, -12.42]
## # spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]
##   id          date        prcp  tmax  tmin
##   <chr>       <date>     <dbl> <dbl> <dbl>
## 1 ASN00009021 2020-01-01     0  31.9  15.3
## 2 ASN00009021 2020-01-02     0  24.9  16.4
## 3 ASN00009021 2020-01-03     6  23.2  13  
## 4 ASN00009021 2020-01-04     0  28.4  12.4
## 5 ASN00009021 2020-01-05     0  35.3  11.6
## # ... with 1,825 more rows
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# example with a tsibble created from climate\_flat}
\NormalTok{raw }\OtherTok{\textless{}{-}}\NormalTok{ climate\_flat }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
\NormalTok{  tsibble}\SpecialCharTok{::}\FunctionTok{as\_tsibble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date)}

\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{  raw }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
\NormalTok{  cubble}\SpecialCharTok{::}\FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}
\NormalTok{dt}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # cubble:   id [5]: nested form
## # bbox:     [115.97, -32.94, 133.55, -12.42]
## # temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
##   id            lat  long  elev name           wmo_id ts                
##   <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
## 1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_ts [366 x 4]>
## 2 ASN00010311 -31.9  117. 179   york            94623 <tbl_ts [366 x 4]>
## 3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_ts [366 x 4]>
## 4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_ts [366 x 4]>
## 5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_ts [366 x 4]>
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# add station{-}based features in the nested form.}
\NormalTok{dt }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{mutate}\NormalTok{(fabletools}\SpecialCharTok{::}\FunctionTok{features}\NormalTok{(ts, tmax, }\FunctionTok{list}\NormalTok{(}\AttributeTok{tmax\_mean =}\NormalTok{ mean, }\AttributeTok{tmax\_var =}\NormalTok{ var)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # cubble:   id [5]: nested form
## # bbox:     [115.97, -32.94, 133.55, -12.42]
## # temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
##   id            lat  long  elev name          wmo_id ts       tmax_mean tmax_var
##   <chr>       <dbl> <dbl> <dbl> <chr>          <dbl> <list>       <dbl>    <dbl>
## 1 ASN00009021 -31.9  116.  15.4 perth airport  94610 <tbl_ts>      25.7    38.6 
## 2 ASN00010311 -31.9  117. 179   york           94623 <tbl_ts>      26.2    51.1 
## 3 ASN00010614 -32.9  117. 338   narrogin       94627 <tbl_ts>      23.7    45.4 
## 4 ASN00014015 -12.4  131.  30.4 darwin airpo~  94120 <tbl_ts>      33.1     3.02
## 5 ASN00015131 -17.6  134. 220   elliott        94236 <tbl_ts>      34.6    24.7
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# read in the .nc file as a ncdf4 class}
\NormalTok{raw }\OtherTok{\textless{}{-}}\NormalTok{ ncdf4}\SpecialCharTok{::}\FunctionTok{nc\_open}\NormalTok{(here}\SpecialCharTok{::}\FunctionTok{here}\NormalTok{(}\StringTok{"data/era5{-}pressure.nc"}\NormalTok{))}
